---
title: "Assignment 5: Data Visualization"
author: "Hanbin Lyu"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 
1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1
#load packages
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

#check working directory
getwd()

#read cvs files
lake = read.csv(
  "Data/Processed/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
  stringsAsFactors = TRUE)
litter = read.csv(
  "Data/Processed/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv",
  stringsAsFactors = TRUE)

#2 
# Check the structure of the data set, make sure dates are read as date format
str(lake)
str(litter)
```

## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
library(ggplot2)

#define my own theme
my_theme = theme(
    line = element_line(color = "darkseagreen4", linewidth = 2,
                        linetype = "solid", lineend = "round"),
    rect = element_rect(fill = "honeydew",color = "darkolivegreen4",
                        linewidth = 1, linetype = "solid"),
    text = element_text(family = "serif", face = "plain",
                        color = "darkolivegreen",size = 12, hjust = 0.5,
                        vjust = 0.5, angle = 0, lineheight = 1.5),
    
    # Modified inheritance structure of text element
    plot.title = element_text(family = "serif", face = "bold",
                              color = "darkolivegreen", size = 16, hjust = 0.5,
                              vjust = 0.5, angle = 0, lineheight = 1.5),
    axis.title.x = element_text(family = "serif", face = "plain",
                              color = "darkolivegreen", size = 12, hjust = 0.5,
                              vjust = 0.5, angle = 0, lineheight = 1.5),
    axis.title.y = element_text(family = "serif", face = "plain",
                              color = "darkolivegreen", size = 12, hjust = 0.5,
                              vjust = 0.5, angle = 0, lineheight = 1.5),
    axis.text = element_text(family = "serif", face = "plain",
                             color = "darkolivegreen", size = 12, hjust = 0.5,
                             vjust = 0.5, angle = 0, lineheight = 1.5),
    
    # Modified inheritance structure of line element
    axis.ticks = element_blank(),
    panel.grid.major =  element_line(color = "gray85", linewidth = 0.5,
                                     linetype = "solid", lineend = "square"),
    panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    plot.background = element_rect(fill = "ivory"),
    panel.background = element_rect(fill = "lightyellow"),
    legend.key = element_rect(fill = "honeydew", linewidth = 0.5),
    
    # Modifiying legend.position
    legend.position = 'right',
    
    complete = TRUE
    )
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4
#check the distribution of data
summary(lake$po4)
summary(lake$tp_ug)

#find the extreme value by looking at the plot
ggplot(lake, aes(x = po4, y = tp_ug, color = lakename,)) +
  #set x and y axis, separate lakes with different color
  geom_point(size = 0.5) +
  #add points for each lake
  geom_smooth(method = "lm", se = FALSE) +
  #add a best fit line
  labs(title = "Total Phosphorus by Phosphate",
       x = "Phosphate",
       y = "Total Phosphorus",
       color = "Lakes") +
  #add title, axis titles, and legend
  scale_color_manual(values = c("Peter Lake" = "cadetblue2", "Paul Lake" = "rosybrown2")) +
   my_theme

#get ride of extreme value
ggplot(lake, aes(x = po4, y = tp_ug, color = lakename,)) +
  #set x and y axis, separate lakes with different color
  geom_point(size = 0.5) +
  #add points for each lake
  geom_smooth(method = "lm", se = FALSE) +
  #add a best fit line
  xlim(-0.233, 50) +  
  #adjust x-axis limits to hide extreme values
  labs(title = "Total Phosphorus by Phosphate",
       x = "Phosphate",
       y = "Total Phosphorus",
       color = "Lakes") +
  #add title, axis titles, and legend
  scale_color_manual(values = c("Peter Lake" = "cadetblue2", "Paul Lake" = "rosybrown2")) +
   my_theme
```

5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5
#make box plot of temperature by month
temperature = ggplot(lake, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  labs(title = "Temp by Month", x = "Months", y = "Temperature", color = "Lakes") +
  scale_color_manual(values = c("Peter Lake" = "cadetblue2", "Paul Lake" = "rosybrown2")) +
  my_theme +
  theme(axis.title = element_blank(), legend.position = "none")
temperature

#make box plot of TP by month
TP = ggplot(lake, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  labs(title = "TP by Month", x = "Months", y = "TP", color = "Lakes") +
  scale_color_manual(values = c("Peter Lake" = "cadetblue2", "Paul Lake" = "rosybrown2")) +
  my_theme +
  theme(axis.title = element_blank(), legend.position = "none")
TP

#make box plot of TN by month
TN = ggplot(lake, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  labs(title = "TN by Month", x = "Months", y = "TN", color = "Lakes") +
  scale_color_manual(values = c("Peter Lake" = "cadetblue2", "Paul Lake" = "rosybrown2")) +
  my_theme +
  theme(axis.title = element_blank(), legend.position = "none")
TN

#combine 3 plots
cowplot = plot_grid(temperature, TP, TN, ncol = 3, align = 'v', rel_widths = c(1.2, 1, 1))
cowplot
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: peter lake has higher TP and TN values overall.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
#filter needles functional group
needles = litter %>% 
  filter(functionalGroup == "Needles")

#plot dry mass of needles by date with NLCD class as color
ggplot(needles, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  geom_line() +
  labs(title = "Dry Mass of Needle Litter by Date",
    x = "Date",
    y = "Dry Mass",
    color = "NLCD Class") +
  scale_color_manual(
    values = c("evergreenForest" = "cadetblue2", "grasslandHerbaceous" = "rosybrown2", "shrubScrub" = "mediumpurple1"),
    labels = c("evergreenForest" = "Evergreen Forest", "grasslandHerbaceous" = "Grassland Herbaceous", "shrubScrub" = "Shrub Scrub")) +
  my_theme +
  theme(axis.text.x = element_text(angle = 60))


#7
#NLCD classes separated into three facets
ggplot(needles, aes(x = collectDate, y = dryMass)) +
  geom_point(color = "darkseagreen3") +
  geom_line(color = "darkseagreen4") +
  facet_wrap(facets = vars(nlcdClass)) +
  labs(title = "Dry Mass of Needle Litter by Date",
    x = "Date",
    y = "Dry Mass") +
  my_theme +
  theme(axis.text.x = element_text(angle = 80, size = 6))
```

Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think facet plot is more effcient because each NLCD class has its own facet, making it easier to focus on individual patterns without clutter. In the plot 6, points are overlapped so it is difficult to distinguish between the different classes, even with distinct colors.